---
title: "Discovering Patterns in Public Health Data"
author: "Adrian G. Zucco"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    theme: cosmo
    self-contained: true
execute:
  warning: false
  message: false
---

## Introduction

Welcome to this hands-on workshop on **pattern recognition in public health
data**. You will work with three synthetic (artificially generated) datasets that
describe individuals, their families, and the locations where they live. The goal
is to discover hidden structure in these data using a combination of:
 
1. **Exploratory Data Analysis (EDA)** -- looking at distributions and
   relationships between variables.
2. **Unsupervised learning** -- grouping similar individuals together *without*
   any predefined labels (clustering).
3. **Supervised learning** -- training a model to *predict* an outcome (sleep
   problems) from the other variables.
4. **Model explainability** -- understanding *how* a complex model makes the
   predictions it does.

No prior R experience is assumed. Each code chunk can be run one at a time;
explanations before and after every chunk describe what is happening and why.


## Setup: install and load packages {#sec-setup}

### Installing packages

The code below checks which packages are missing from your system and installs
only those. You only need to run this **once** (or whenever you work on a new
machine). If all packages are already installed the chunk does nothing.

```{r}
#| label: install-packages
#| eval: false

# List every package used in this notebook
required_packages <- c(
  "here",        # Robust file paths that work regardless of working directory
  "skimr",       # Compact summary statistics
  "tidyverse",   # Data wrangling and visualisation (dplyr, ggplot2, readr …)
  "tidymodels",  # Modelling framework (recipes, rsample, parsnip, yardstick …)
  "corrplot",    # Correlation matrix plots
  "ranger",      # Fast random-forest implementation
  "lightgbm",    # Gradient-boosted decision trees
  "bonsai",      # tidymodels integration for LightGBM
  "shapviz",     # SHAP visualisations (LightGBM computes SHAP values natively)
  "GGally",      # Extended ggplot2 (pair plots)
  "umap",        # UMAP dimensionality reduction
  "dbscan",      # HDBSCAN density-based clustering
  "viridis",     # Colour-blind-friendly palettes
  "ggrepel"      # Non-overlapping text labels in plots
)

# Install only the packages you don't already have
missing <- required_packages[!required_packages %in% installed.packages()[, "Package"]]
if (length(missing) > 0) {
  message("Installing missing packages: ", paste(missing, collapse = ", "))
  install.packages(missing)
}
```

::: {.callout-tip}
## Trouble installing packages?
If `install.packages()` fails for a package, try installing the
[pak](https://pak.r-lib.org/) helper first (`install.packages("pak")`) and
then run `pak::pkg_install("package_name")`. `pak` is better at resolving
complex dependencies.
:::

### Loading packages

Once installed, we **load** every package into our R session.

```{r}
#| label: load-packages

library(here)         # Always-correct file paths
library(tidyverse)    # Data manipulation and visualisation
library(skimr)        # Summary statistics
library(tidymodels)   # Modelling framework
library(corrplot)     # Correlation plots
library(ranger)       # Random forests
library(shapviz)      # SHAP visualisation (uses LightGBM's native SHAP engine)
library(GGally)       # Pair plots
library(umap)         # UMAP
library(dbscan)       # HDBSCAN
library(viridis)      # Colour palettes
library(ggrepel)      # Label repulsion
library(lightgbm)     # LightGBM engine
library(bonsai)       # tidymodels <-> LightGBM bridge
```

### Setting the project root

The `here` package figures out where the **project root** is (the folder that
contains the `.Rproj` file or a `.here` file). Every file path in this notebook
uses `here()` so it works no matter which folder you opened RStudio from.

```{r}
#| label: check-root

# This should print the top-level folder of the workshop repository
here::here()
```

::: {.callout-note}
## Why `here()`?
A very common error when starting with R is
*"cannot open file '../data/individuals.csv': No such file or directory"*.
This happens when your *working directory* is not where R expects it to be.
`here()` solves this by always building paths relative to the project root,
regardless of where you run the code from.
:::


## Data loading and merging

We have three CSV files in the `data/` folder:

| File                | What it contains |
|---------------------|------------------|
| `individuals.csv`   | One row per person: age, sex, occupation, lifestyle and health variables |
| `locations.csv`     | One row per location: population density, average income and wealth |
| `families.csv`      | Links each person to a family and that family's income |

```{r}
#| label: load-data

df_ind <- read_csv(here("data", "individuals.csv"))
df_loc <- read_csv(here("data", "locations.csv"))
df_fam <- read_csv(here("data", "families.csv"))
```

We merge (join) these three tables into a single data frame so that every row
contains *all* information about one individual: their personal attributes, the
characteristics of the location they live in, and their family income.

```{r}
#| label: merge-data

df <- df_ind |>
  left_join(df_loc, by = "Location") |>
  left_join(df_fam, by = "Person ID")
```

A `left_join` keeps every row from the left table (`df_ind`) and attaches
matching columns from the right table. If a person's location or family ID does
not appear in the lookup table the cells will be filled with `NA`.

Let us take a first look at the merged data. `skim()` gives a rich summary
with counts of missing values, means, histograms, etc.

> **Question for reflection:** How many rows and columns does the merged dataset
> have? Are there any missing values?

```{r}
#| label: inspect-data

glimpse(df)
skim(df)
```


## Exploratory Data Analysis (EDA)

Before building any model it is crucial to *look* at the data. We use **pair
plots** to visualise distributions (on the diagonal) and bivariate relationships
(off the diagonal) for each dataset.

### Pair plots

> **What to look for:**
>
> - Are some variables clearly correlated (e.g. a rising cloud of points)?
> - Do any distributions look skewed or have unusual peaks?
> - Do categorical variables (like *Sex* or *Occupation*) show different patterns
>   for different groups?

```{r}
#| label: pairplot-individuals
#| fig-width: 10
#| fig-height: 10

GGally::ggpairs(
  df_ind,
  progress = FALSE,
  title = "Pairplot of Individual-Level Features"
)
```

```{r}
#| label: pairplot-locations
#| fig-width: 8
#| fig-height: 8

GGally::ggpairs(
  df_loc,
  progress = FALSE,
  title = "Pairplot of Location-Level Features"
)
```

### Unique values in categorical columns

```{r}
#| label: unique-values

df |>
  select(where(is.character)) |>
  map(unique)
```


## Data preprocessing: encoding

Analyses using computational models require us to map our data and concepts into
numbers in a way that we can represent meaningful differences between observations.
We therefore need to convert categorical variables (like *Sex* or *Occupation*)
into numeric columns. A common approach is **one-hot encoding**: each category
becomes its own 0/1 column. This is sometimes done under the hood when using
statistical models, but here we do it explicitly so we can make sure that
different categories are numerically differentiated.

We use a tidymodels **recipe** -- a reusable set of preprocessing instructions:

```{r}
#| label: encode-data

rec <- recipe(`Sleep problems` ~ ., data = df) |>
  # Make the target a factor so tidymodels treats this as classification
  step_mutate(`Sleep problems` = factor(`Sleep problems`)) |>
  # Remove ID columns and the raw Location name (its features are already merged in)
  step_rm(`Person ID`, `Family ID`, Location) |>
  # One-hot encode every remaining categorical predictor;
  # one_hot = FALSE (default) automatically drops one level per variable
  # to avoid multicollinearity -- equivalent to drop_first=True in Python
  step_dummy(all_nominal_predictors(), one_hot = FALSE)

# prep() learns the encoding from the data; bake() applies it
data_encoded <- rec |>
  prep() |>
  bake(new_data = NULL)
```

::: {.callout-tip}
## What is multicollinearity?
If we keep *both* `Sex_F` and `Sex_M` columns, one is a perfect linear function
of the other (`Sex_M = 1 - Sex_F`). Some algorithms struggle with such
redundancy, so we drop one. Using `one_hot = FALSE` in `step_dummy()` handles
this automatically by dropping the first level of each categorical variable.
:::

```{r}
#| label: inspect-encoded

skim(data_encoded)
glimpse(data_encoded)
```


## Visualising the encoded data

### Feature heatmap

A heatmap lets us see all **1 000 individuals x features** at a glance. We
*scale* every column to zero mean and unit variance first so that features with
different units (e.g. age in years vs. BMI) are visually comparable.

```{r}
#| label: heatmap
#| fig-width: 10
#| fig-height: 8

data_encoded |>
  select(-`Sleep problems`) |>
  scale() |>
  as.matrix() |>
  heatmap(
    Colv = NA, Rowv = NA,
    col  = viridis::viridis(100),
    main = "Scaled Feature Heatmap",
    margins = c(5, 10)
  )
```

### Correlation plot

Next we look at **Spearman correlations** between features. This metric is
robust to non-linear relationships and outliers.

> **What to look for:** Clusters of strongly correlated features may represent
> related concepts or potential *interactions*. However, such correlations
> can be overinterpreted, they do not necessarily imply a causal relationship, and may
> be driven by confounding variables or data artifacts.


## Unsupervised learning: clustering {#sec-clustering}

Clustering tries to find **natural groupings** in the data without using any
outcome variable. We will try two very different approaches:

| Method | Key idea |
|--------|----------|
| **K-means** | Partition the data into exactly *k* spherical clusters of similar size |
| **HDBSCAN** | Find clusters of *varying* shape and density; can label points as "noise" |

```{r}
#| label: seed-clustering

set.seed(2026) # Ensures results are reproducible
features_for_clustering <- data_encoded |> select(-`Sleep problems`)
```

### K-means clustering

K-means needs us to choose the number of clusters in advance. Here we set
`k = 3` as a starting point. 

> Assessing how many clusters to choose is one of the biggest challenges in unsupervised learning.
> In practice one would try to assess by different methods such as an **elbow plot** or
> **silhouette score** to pick *k*.

```{r}
#| label: kmeans

km_model <- kmeans(
  scale(features_for_clustering), # Scaling is critical -- otherwise features
  centers = 3,                    #   measured in large units dominate the distance
  nstart  = 25                    # Try 25 random starting configurations
)
```

#### Visualising clusters with PCA

Our data has many columns. To *plot* the clusters in 2-D we project them onto the
first two **principal components (PCs)** -- the two directions that capture the
most variance.

```{r}
#| label: pca

pca_res_km <- prcomp(features_for_clustering, scale. = TRUE)

pca_variance <- pca_res_km$sdev^2
prop_var      <- pca_variance / sum(pca_variance)
cum_var       <- cumsum(prop_var)

tibble(
  PC = seq_along(prop_var),
  `Variance explained` = prop_var,
  `Cumulative variance` = cum_var
) |>
  head(10) |>
  print()
```

Now we overlay the K-means cluster labels and the **loading arrows** (which show
how each original feature contributes to the principal components):

```{r}
#| label: kmeans-pca-plot
#| fig-width: 10
#| fig-height: 8

# PC scores + cluster labels
pca_df_km <- as_tibble(pca_res_km$x[, 1:2]) |>
  mutate(cluster = factor(km_model$cluster))

# Loading vectors
loadings_data <- as_tibble(pca_res_km$rotation[, 1:2], rownames = "Variable") |>
  rename(PC1_loading = PC1, PC2_loading = PC2)

# Axis labels showing % variance explained
pc1_lab <- sprintf("PC1 (%.1f%% variance)", prop_var[1] * 100)
pc2_lab <- sprintf("PC2 (%.1f%% variance)", prop_var[2] * 100)

# Scale loading arrows so they are visible on the score plot
plot_diag <- sqrt(diff(range(pca_df_km$PC1))^2 + diff(range(pca_df_km$PC2))^2)
max_load  <- max(sqrt(loadings_data$PC1_loading^2 + loadings_data$PC2_loading^2))
scale_f   <- if (max_load > 1e-6) (plot_diag * 0.35) / max_load else 1

ggplot(pca_df_km, aes(x = PC1, y = PC2)) +
  geom_point(aes(colour = cluster), alpha = 0.6, size = 2) +
  geom_segment(
    data = loadings_data,
    aes(x = 0, y = 0,
        xend = PC1_loading * scale_f,
        yend = PC2_loading * scale_f),
    arrow     = arrow(length = unit(0.2, "cm")),
    colour    = "grey20",
    linewidth = 0.6,
    inherit.aes = FALSE
  ) +
  ggrepel::geom_text_repel(
    data = loadings_data,
    aes(x = PC1_loading * scale_f * 1.1,
        y = PC2_loading * scale_f * 1.1,
        label = Variable),
    colour = "grey10", size = 3,
    inherit.aes = FALSE,
    max.overlaps = Inf
  ) +
  labs(
    title  = "K-means Clusters with PCA Loadings (PC1 vs PC2)",
    x      = pc1_lab,
    y      = pc2_lab,
    colour = "Cluster"
  ) +
  theme_minimal(base_size = 12) +
  coord_fixed() +
  scale_colour_viridis_d(option = "C")
```

> **How to read this plot:**
>
> - Each **point** is a person, coloured by their cluster.
> - The **arrows** show which features drive the separation. Longer arrows =
>   more influential features. Arrows pointing in the same direction are
>   positively correlated on these components.
> - `coord_fixed()` makes the aspect ratio 1:1, which is important so distances
>   are not visually distorted.


### HDBSCAN clustering

HDBSCAN does not require us to choose the number of clusters. Instead, it finds
dense regions and labels sparse points as **noise** (cluster 0).

Before running HDBSCAN we reduce dimensionality with **UMAP** (Uniform Manifold
Approximation and Projection), which preserves local neighbourhood structure
better than PCA and works well as input for density-based clustering.

::: {.callout-note}
## Why UMAP before HDBSCAN?
Density-based methods like HDBSCAN struggle in high-dimensional spaces — a
phenomenon known as the **curse of dimensionality**. As the number of dimensions
grows, distances between points become increasingly uniform, making it hard to
distinguish "dense" from "sparse" regions. UMAP compresses the data into a
low-dimensional (2-D) embedding while preserving local neighbourhood structure,
giving HDBSCAN a cleaner, more meaningful distance landscape to work with.
:::

```{r}
#| label: umap-hdbscan
#| fig-width: 9
#| fig-height: 7

# UMAP reduction to 2-D
umap_res <- umap::umap(
  scale(features_for_clustering),
  n_neighbors  = 15,
  min_dist     = 0.1,
  random_state = 2025
)

umap_df <- as_tibble(umap_res$layout, .name_repair = "minimal") |>
  setNames(c("UMAP1", "UMAP2"))

# HDBSCAN -- minPts controls the minimum cluster size
hdb_clusters <- dbscan::hdbscan(umap_df, minPts = 10)
umap_df$cluster <- factor(hdb_clusters$cluster)

cat("Clusters found (excl. noise):",
    length(unique(hdb_clusters$cluster[hdb_clusters$cluster != 0])), "\n")
cat("Noise points (cluster 0):", sum(hdb_clusters$cluster == 0), "\n")

ggplot(umap_df, aes(x = UMAP1, y = UMAP2, colour = cluster)) +
  geom_point(size = 1.5, alpha = 0.7) +
  labs(
    title  = "HDBSCAN Clusters on UMAP Embedding",
    x      = "UMAP Dimension 1",
    y      = "UMAP Dimension 2",
    colour = "Cluster"
  ) +
  theme_minimal() +
  scale_colour_viridis_d(option = "D")
```

> **K-means vs HDBSCAN:** K-means always assigns every point to a cluster and
> assumes roughly spherical clusters of similar size. HDBSCAN can discover
> irregularly shaped clusters and exposes when a point does not belong
> anywhere (noise). Try changing `minPts` and see how the results change!

### Summary statistics by cluster

To **interpret** the clusters we go back to the original (unencoded) variables
and compute summary statistics within each cluster.

> **Question for reflection:** Do the clusters correspond to recognisable
> subpopulations? For example, do some clusters have higher stress, lower
> physical activity, or more sleep problems?

```{r}
#| label: cluster-summaries

df_with_clusters <- df |>
  mutate(cluster_hdbscan = factor(hdb_clusters$cluster))

valid_clusters <- df_with_clusters |>
  filter(cluster_hdbscan != 0)

if (nrow(valid_clusters) > 0) {
  top_ids <- valid_clusters |>
    count(cluster_hdbscan, sort = TRUE) |>
    slice_head(n = 5) |>
    pull(cluster_hdbscan)

  for (cid in top_ids) {
    cat("\n--- HDBSCAN Cluster", as.character(cid),
        "(n =", sum(valid_clusters$cluster_hdbscan == cid), ") ---\n")
    valid_clusters |>
      filter(cluster_hdbscan == cid) |>
      select(-cluster_hdbscan) |>
      skim() |>
      print()
  }
} else {
  cat("No valid (non-noise) HDBSCAN clusters found.\n")
}
```


## Supervised learning: predicting sleep problems {#sec-supervised}

So far we have explored the data without a specific labels/outcome. Now we switch to
**supervised learning**: we ask the model to predict whether an individual has
**sleep problems** (0 = no, 1 = yes) from all other features.

### Train/test split

We split the data randomly into a **training set** (80 %) used to fit the model and a
**test set** (20 %) used to evaluate it. Stratifying on the outcome ensures that
the proportion of sleep problems is similar in both sets.

```{r}
#| label: train-test-split

set.seed(2025)
split      <- initial_split(data_encoded, prop = 0.80, strata = `Sleep problems`)
train_data <- training(split)
test_data  <- testing(split)

cat("Training set:", nrow(train_data), "rows\n")
cat("Test set:    ", nrow(test_data), "rows\n")
```

### Training a LightGBM model

[LightGBM](https://lightgbm.readthedocs.io/) is a gradient-boosting framework
that builds an ensemble of decision trees sequentially, each one correcting the
errors of the previous one. It is fast, handles missing values natively, and
works well with tabular data.
We use the `bonsai` package so that LightGBM plugs neatly into the tidymodels workflow.

::: {.callout-note}
## LightGBM and categorical features
One-hot encoding is a fundamental concept worth understanding, but advanced
tree-based models like LightGBM can natively handle raw categorical columns
without any manual encoding. In a production workflow you could pass `Occupation`
and `Sex` directly and use the `categorical_feature` engine argument, letting the
model find the optimal splits internally.
:::

```{r}
#| label: train-lgbm

lgbm_spec <- boost_tree(
  mode  = "classification",
  trees = 100  # Number of boosting rounds
) |>
  set_engine("lightgbm", seed = 2025)

wf <- workflow() |>
  add_model(lgbm_spec) |>
  add_formula(`Sleep problems` ~ .)

lgbm_fit <- wf |> fit(data = train_data)
```

### Evaluating the model

We predict on the held-out test set and compute several **classification
metrics**:

| Metric | What it tells you |
|--------|-------------------|
| **Accuracy** | Overall fraction of correct predictions |
| **ROC AUC** | How well the model ranks positive cases above negative ones (1 = perfect) |
| **Sensitivity** (recall) | Fraction of actual positives correctly identified |
| **Specificity** | Fraction of actual negatives correctly identified |
| **PR AUC** | Area under the precision-recall curve (useful with imbalanced classes, aka low number of cases) |

> **Question for reflection:** Is accuracy alone a good metric here? What if
> only 10 % of people have sleep problems -- could a model that always predicts
> "no" still reach 90 % accuracy?

```{r}
#| label: evaluate-lgbm

preds_df <- lgbm_fit |>
  predict(test_data, type = "prob") |>
  bind_cols(predict(lgbm_fit, test_data)) |>
  bind_cols(test_data |> select(`Sleep problems`))

positive_level <- levels(train_data$`Sleep problems`)[2]
prob_col       <- paste0(".pred_", positive_level)

# Majority-class baseline accuracy (context for the model's accuracy)
baseline_acc <- max(mean(train_data$`Sleep problems` == levels(train_data$`Sleep problems`)[1]),
                    mean(train_data$`Sleep problems` == levels(train_data$`Sleep problems`)[2]))
cat(sprintf("Baseline Accuracy (majority class): %.3f\n", baseline_acc))

classification_metrics <- metric_set(
  accuracy, roc_auc, sensitivity, specificity, pr_auc
)

eval_results <- preds_df |>
  classification_metrics(
    truth       = `Sleep problems`,
    estimate    = .pred_class,
    !!sym(prob_col),
    event_level = "second"
  )

print(eval_results)
```

### Confusion matrix

The **confusion matrix** shows the counts of true positives, true negatives,
false positives and false negatives in a 2x2 table.

```{r}
#| label: confusion-matrix
#| fig-width: 6
#| fig-height: 5

conf_mat_res <- conf_mat(preds_df, truth = `Sleep problems`, estimate = .pred_class)
print(conf_mat_res)

autoplot(conf_mat_res, type = "heatmap") +
  labs(title = "Confusion Matrix -- Sleep Problems (LightGBM)") +
  theme_minimal()
```


## A linear baseline: logistic regression {#sec-logistic}

Before diving into SHAP, we train a **logistic regression** as a linear
baseline. This model assumes that risk factors combine *additively* and *linearly* on the log-odds
scale. In a truly complex system — where variables interact non-linearly and
effects depend on context — that assumption may not hold. Comparing the linear
model against LightGBM reveals how much is lost (or gained) by relaxing the
linearity constraint.

We add a `step_normalize()` to the recipe so that all coefficients are on a
comparable scale: each one represents the change in log-odds per **standard
deviation** of the feature.

```{r}
#| label: train-lr

lr_rec <- recipe(`Sleep problems` ~ ., data = train_data)

lr_spec <- logistic_reg() |>
  set_engine("glm")

lr_wf <- workflow() |>
  add_recipe(lr_rec) |>
  add_model(lr_spec)

lr_fit <- lr_wf |> fit(data = train_data)
```

### Head-to-head performance comparison

The table below puts both models side by side on the same test set.

> **Question for reflection:** How much does the non-linear LightGBM model
> outperform the linear logistic regression? A large gap suggests the data
> contains important **interactions or non-linearities** that a linear model
> cannot capture — a hallmark of complex systems. A small gap suggests the
> relationship is largely additive and linear.

```{r}
#| label: compare-models

lr_preds_df <- lr_fit |>
  predict(test_data, type = "prob") |>
  bind_cols(predict(lr_fit, test_data)) |>
  bind_cols(test_data |> select(`Sleep problems`))

lr_prob_col <- paste0(".pred_", positive_level)

lr_eval <- lr_preds_df |>
  metric_set(accuracy, roc_auc, pr_auc)(
    truth       = `Sleep problems`,
    estimate    = .pred_class,
    !!sym(lr_prob_col),
    event_level = "second"
  ) |>
  mutate(model = "Logistic Regression")

lgbm_eval_sub <- eval_results |>
  filter(.metric %in% c("accuracy", "roc_auc", "pr_auc")) |>
  mutate(model = "LightGBM")

comparison_df <- bind_rows(lr_eval, lgbm_eval_sub) |>
  select(model, .metric, .estimate) |>
  tidyr::pivot_wider(names_from = .metric, values_from = .estimate)

print(comparison_df)
```

## Model explainability: SHAP values {#sec-shap}

A model that predicts well could be useful but more interesting is to know
what the model learned from the data. Traditional statistical models (like logistic regression)
have coefficients that are based on the mean effect of each feature across the whole dataset.
SHAP (SHapley Additive exPlanations) can help us decompose the contributions of each feature,
and for each individual prediction, based on the idea of Shapley values from cooperative game theory.

- A **positive** SHAP value for a feature means it *pushes the prediction
  towards* sleep problems for that person.
- A **negative** SHAP value means it *pushes the prediction away from* sleep
  problems.

### Computing SHAP values

```{r}
#| label: compute-shap

actual_lgbm_model <- lgbm_fit |>
  extract_fit_parsnip() |>
  (\(x) x$fit)()  # Extract the underlying lgb.Booster object

X_pred_shap <- test_data |>
  select(-`Sleep problems`) |>
  as.data.frame()

sv <- shapviz::shapviz(
  actual_lgbm_model,
  X_pred = data.matrix(X_pred_shap),
  X      = X_pred_shap
)
```

### Global feature importance

The **bar plot** shows the mean absolute SHAP value for each feature -- a measure
of that feature's overall importance across all individuals in the test set.

```{r}
#| label: shap-bar
#| fig-width: 8
#| fig-height: 6

sv_importance(sv, kind = "bar", max_display = 15) +
  ggtitle("Global Feature Importance (Mean |SHAP|)") +
  theme_minimal(base_size = 10)
```

### SHAP beeswarm plot

The **beeswarm** plot goes one step further. Each dot is one individual. The
horizontal position shows the SHAP value (how much the feature pushed the
prediction) and the colour shows the feature's *actual value* (red = high,
blue = low).

```{r}
#| label: shap-beeswarm
#| fig-width: 8
#| fig-height: 6

sv_importance(sv, kind = "beeswarm", max_display = 15) +
  ggtitle("SHAP Summary (Beeswarm) -- LightGBM") +
  theme_minimal(base_size = 10)
```

> **How to read the beeswarm:** If a feature shows red dots on the right and
> blue dots on the left, then *higher values of that feature increase the
> predicted risk of sleep problems*. The reverse pattern means higher values are
> protective.

### Logistic regression coefficients vs LightGBM SHAP values

Now we can directly compare what the two models have learned about feature
importance.

- **LR coefficients** (standardised) tell us the *average*, *linear* log-odds
  contribution of each feature, assuming all others are held constant.
- **Mean |SHAP|** values tell us the *average magnitude* of each feature's
  contribution to the LightGBM predictions — capturing non-linear and interaction
  effects as well.

Features that rank similarly in both views are robustly important regardless of
modelling assumptions. Features that rank very differently hint at
**non-linearities or interactions** that the linear model misses — a key
signature of complex systems.

```{r}
#| label: shap-vs-lr
#| fig-width: 13
#| fig-height: 7

# Mean absolute SHAP per feature (LightGBM)
shap_imp <- colMeans(abs(sv$S)) |>
  tibble::enframe(name = "term", value = "mean_abs_shap")

# Standardised LR coefficients (absolute value for magnitude)
lr_coefs <- lr_fit |>
  broom::tidy() |>
  filter(term != "(Intercept)") |>
  mutate(
    term = stringr::str_replace_all(term, "`", ""),
    abs_coef = abs(estimate)
  ) |>
  select(term, abs_coef)

# Join and sort by SHAP importance
compare_imp <- shap_imp |>
  left_join(lr_coefs, by = "term") |>
  tidyr::replace_na(list(abs_coef = 0)) |>
  arrange(desc(mean_abs_shap))

# Show the table
print(compare_imp)
```

> **What to look for:**
>
> - Features near the **top of both panels** are important under both linear and
>   non-linear assumptions — these are your most reliable signals.
> - Features that appear only in the **SHAP panel** (high SHAP, low LR
>   coefficient) are likely involved in **interactions or threshold effects** that
>   the logistic model cannot represent.
> - Features that appear only in the **LR panel** may have a consistent additive
>   effect that the tree model spreads across correlated features.

### SHAP dependence plots

A dependence plot zooms in on **one feature** and shows exactly how its value
relates to its SHAP contribution. A second feature is used for colour to reveal
**interactions**.

> **Try it yourself:** Replace `"Physical Activity Level"` or `"Stress Levels"`
> with another feature name (run `colnames(X_pred_shap)` to see all options) and
> explore different interactions.

```{r}
#| label: shap-dependence
#| fig-width: 8
#| fig-height: 5

# Helper: safely create a dependence plot, skipping if a column is missing
plot_shap_dep <- function(sv_obj, feature, color_var = NULL, data_cols) {
  if (!(feature %in% data_cols)) {
    cat("Skipping:", feature, "not found in prediction data.\n")
    return(invisible(NULL))
  }
  if (!is.null(color_var) && !(color_var %in% data_cols)) {
    cat("Warning:", color_var, "not found; plotting without colour.\n")
    color_var <- NULL
  }
  title <- paste0("SHAP Dependence: ", feature)
  if (!is.null(color_var)) title <- paste0(title, " (colour: ", color_var, ")")

  sv_dependence(sv_obj, v = feature, color_var = color_var) +
    ggtitle(title) +
    theme_minimal(base_size = 10)
}

plot_shap_dep(sv, "Physical Activity Level", "Age", colnames(X_pred_shap))
plot_shap_dep(sv, "Stress Levels", "Age", colnames(X_pred_shap))
```


## Wrapping up

In this workshop you have:

- **Loaded and merged** three related datasets.
- **Explored** variable distributions and correlations.
- **Discovered subgroups** using K-means and HDBSCAN clustering.
- **Predicted sleep problems** with a LightGBM model.
- **Benchmarked** it against a logistic regression and compared their
  performance and what each model "learned" about feature importance.
- **Explained** the model's predictions using SHAP values.

### Ideas for further exploration

- Change the number of clusters in K-means (`centers = ?`) or the density
  threshold in HDBSCAN (`minPts = ?`). How do the results change?
- Try predicting a different outcome (e.g. `Depressive symptoms` instead of
  `Sleep problems`).
- Add or remove features from the model -- does performance improve?
- Look at individual SHAP values for specific people in the test set to
  understand *their* prediction.